---
title: "HW06: Income Data"
author: "Julia Du"
date: "2/18/2021"
output: 
  github_document:
    toc: true
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

### Load necessary libraries

```{r, echo = TRUE}
library(reprex)
library(tidyverse)
library(glue)
library(knitr)
library(lubridate)

theme_set(theme_minimal())
```

# Get the Data

I used Tidy Tuesday's [02/09/2021 challenge data](https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-02-09). This particular challenge comes from the Urban Institute and looks at income inequality. I also am using a dataset of presidents and their term length [here](https://gist.github.com/namuol/2657233).

```{r getdata}
# downloading income inequality data
student_debt <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/student_debt.csv')
retirement <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/retirement.csv')
home_owner <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/home_owner.csv')
income_aggregate <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_aggregate.csv')
income_mean <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_mean.csv')

# also download list of presidents & terms
president <- readr::read_csv('https://gist.githubusercontent.com/namuol/2657233/raw/74135b2637e624848c163759be9cd14ae33f5153/presidents.csv')
```

# Simplifying (not tidying) Data

The income data is already tidy, but I am further simplifying some of it for easier analysis. 

Originally, student_debt.csv (i.e. its column loan_debt) and retirement.csv (i.e. its column retirement) are in 2016 dollars, but the income_mean.csv (i.e. its column income_dollars) is in 2019 dollars. For consistency, I'm normalizing all data to 2019 dollars, using [inflation rate info](https://www.calculator.net/inflation-calculator.html?cstartingamount1=1&cinyear1=2016&coutyear1=2019&calctype=1&x=83&y=13) based off average annual CPI data in the U.S. from 1914 to 2020.

I am also simplifying datasets so that possible race values are only white, non-white Hispanic, Black alone, or Asian alone.


```{r simplify}
# rate between 2016 & 2019 dollars
start_2016 <- 1.05940

# normalize dollar amounts to 2019
student_debt <- student_debt %>%
  mutate(loan_debt = loan_debt * start_2016)

retirement <- retirement %>%
  mutate(retirement = retirement *start_2016)

# renaming middle quintile to "Third" for future df joining 
# & keeping only 2019 dollar amounts
income_mean <- income_mean %>%
  filter(dollar_type == "2019 Dollars") %>%
  mutate(income_quintile = if_else(income_quintile == "Middle", "Third", income_quintile))

# reordering income quintile levels & simplify race values
income_leveled <- c("Lowest", "Second", "Third", "Fourth", "Highest", "Top 5%")

level_simple_data <- function(df) {
  df %>%
  mutate(income_quintile = factor(income_quintile, levels = income_leveled)) %>%
  filter(!race %in% c("Asian Alone or in Combination", "Black Alone or in Combination", "White Alone"))
}

income_aggregate <- level_simple_data(income_aggregate) 
income_mean <- level_simple_data(income_mean)
```

# Analyzing the Data

## Exploring Broad Income Trends

```{r}
# income share over time
income_aggregate %>%
  filter(race == "All Races") %>%
  ggplot(mapping = aes(x = year, y = income_share/100, color = income_quintile, group = income_quintile)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
    labs(title = "Income Share of each Income Group over Time",
       x = "Year", y = "Percentage of income share",
       caption = "Source: The Urban Institute", color = "Income Quintile")

# mean income over time
income_mean %>%
  filter(race == "All Races") %>%
  ggplot(mapping = aes(x = year, y = income_dollars, color = income_quintile, group = income_quintile)) +
  geom_line() +
  labs(title = "Mean Income of each Income Group over Time",
       x = "Year", y = "Mean income (2019 dollars)",
       caption = "Source: The Urban Institute", color = "Income Quintile") +
  scale_y_continuous(labels = scales::dollar)
```

Looking at the 1st graph, it’s clear that the highest income quintile has steadily increased its income share over time (from about 42% in the mid-1960s to over 50% in 2019) – as has the top 5% of Americans. Correspondingly, the lower 4 income quintiles have seen their income share decline, lending credence to the fact that there is widening income inequality.
	
The 2nd graph tells a similar story. The bottom 4 income quintiles have seen little-to-no change in their mean income (with the lowest quintile looking particularly stagnant), but the highest quintile has increased from about \$125,000 in the mid-1960s to over \$250,00 in 2019. The top 5% has seen even more dramatic change over the past decades, rising from just under \$200,000 in 1967 to nearly $450,000 in 2019. 


## Exploring Income by Race

```{r}
# mean income for quintile by race
income_mean %>%
  ggplot(mapping = aes(x = income_quintile, y = income_dollars, fill = income_quintile)) +
  geom_violin() +
  geom_boxplot(width = 0.9, color = "grey", alpha = 0.3) +
  facet_wrap(race ~ .) +
    scale_y_continuous(labels = scales::dollar) +
  labs(title = "Mean Income of each Income Group across Race",
       x = "", y = "Mean income (2019 dollars)",
       caption = "Source: The Urban Institute",
       fill = "Income Quintile") +
  theme(axis.text.x = element_blank()) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 1))

# mean income overall by race
income_mean %>%
  filter(income_quintile == "Third") %>%
    group_by(year, race) %>%
  ggplot(mapping = aes(x = year, y = income_dollars, color = race)) +
  geom_point(alpha = .3) + 
  geom_smooth(method = lm, se = FALSE) +
  scale_color_viridis_d(option = "plasma", end = .7) +
  labs(title = "Median Average Income by Race over Time",
       x = "Year", y = "Average income (2019 dollars)",
       caption = "Source: The Urban Institute", color = "Race") +
  scale_y_continuous(labels = scales::dollar)
```

The 1st graph shows that the aforementioned increase in income for top quintiles has not been experienced equally by different racial groups. Asians in the top 5% seem to have a mean income around \$425,000, which is much higher than the typical mean income of \$300,000 for all races in the top 5%. White, non-Hispanic Americans in the top 5% also have above-typical mean incomes, whereas Black and Hispanic Americans are notably lower – Black Americans in the top 5% have a mean income less than half of their Asian counterparts’, about $200,000. The other income quintiles see a similar pattern: Asians in every quintile tend to have a mean income higher than the average for all races, whereas Black Americans have a mean income below that all-race average.

The next graph further confirms this discrepancy. Filtering for the 3rd income quintile to find the actual median average income of the US, we see that Black and Hispanic families have lower mean incomes than the average for all races – and white and Asian families have higher-than-average mean incomes. It is worth noting that there is a smaller amount of data available on Asian families, which may partly influence Asian families’ markedly higher mean incomes. Over time, we do see a general upward trend in mean income for all races in the 3rd income quintile – but race seems to play a significant role in determining mean income.


## Other Metrics of Economic Inequality Across Race

```{r, message = FALSE}
# creating comparable mean income table 
meaninc_compare <- income_mean %>% 
  filter(income_quintile == "Third") %>%
  mutate(race = if_else(race == "Black Alone", "Black", race)) %>%
  mutate(race = ifelse(race == "White, Not Hispanic", "White", race))

meaninc_compare %>%
  inner_join(student_debt, by = c("year", "race")) %>%
  inner_join(retirement, by = c("year", "race")) %>%
  group_by(race) %>%
  summarize(mean(income_dollars), mean(loan_debt), mean(retirement)) %>%
  knitr::kable(
    caption = "Averages of Economic Inequality Metrics (2019 dollars)", 
    col.names = c(
      "Race",
      "Mean income", 
      "Loan debt", 
      "Retirement savings"),
    digits = 2)

compare_inc <- function(df, var){
meaninc_compare %>% 
  inner_join({df}, by = c("year", "race")) %>%
  ggplot(mapping = aes(x = income_dollars, y = {{var}}, color = race)) +
  geom_point(alpha = .3) + 
  geom_smooth(method = lm, se = FALSE) +
    scale_color_viridis_d(end = .7) +
  scale_x_continuous(labels = scales::dollar) +
    theme(legend.position = "bottom")
}

# retirement
compare_inc(retirement, retirement) +
  labs(title = "Median Average Income vs. Average Retirement Savings", 
       subtitle = "(2019 dollars)",
       x = "Average family income", y = "Average family retirement savings",
       caption = "Source: The Urban Institute",
       color = "Race") +
  scale_y_continuous(labels = scales::dollar) 


# student debt
compare_inc(student_debt, loan_debt) +
  labs(title = "Median Average Income vs. Average Student Loan Debt", 
       subtitle = "(2019 dollars)",
       x = "Average family income", y = "Average family student loan debt for aged 25-55",
       caption = "Source: The Urban Institute",
       color = "Race") +
  scale_y_continuous(labels = scales::dollar)

# homeownership
compare_inc(home_owner, home_owner_pct) +
  labs(title = "Median Average Income vs. Home Ownership",
       x = "Average family income (2019 dollars)", y = "Home ownership percentage for families",
       caption = "Source: The Urban Institute",
       color = "Race") +
  scale_y_continuous(labels = scales::percent)
```

As noted above, Black and Hispanic families generally see lower mean incomes than their white counterparts. This pattern continues for other income metrics like retirement savings and home ownership. White families have the highest average mean incomes and retirement savings, but Black families have the highest average student loan debt. White families also tend to have higher retirement savings overall and are much more likely to own their homes (about 70% of white families own their homes, compared to 45% of Black families), but their average student loan debt is similar to Black families’. Hispanic families have notably lower student loan debt, but we caution against hasty conclusions as there is not much data available for either student loan debt or retirement savings. 


## Percentages of Home Ownership & Student Loan Debt

```{r}
home_owner %>% 
  inner_join(student_debt, by = c("year", "race")) %>%
  select(-loan_debt) %>%
  pivot_longer(c(home_owner_pct, loan_debt_pct), 
               names_to = "type", values_to = "percentages") %>%
#  mutate(type = if_else())
  ggplot(mapping = aes(x = race, fill = type)) +
  geom_boxplot(aes(y = percentages)) +  
  labs(title = "Home Ownership and Loan Debt in Families",
       x = "Race", y = "Percentage of families",
       caption = "Source: The Urban Institute",
       fill = "") +
  scale_fill_discrete(labels = c("Home ownership", "Student loan debt")) +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "bottom")

```

This boxplot clarifies what we observed: white families have much higher %s of home ownership, but the % of families with student loan debt is similar to Black families. Hispanic families have similar home ownership percentages as Black families, but are less likely to have student loan debt. For all available races, home ownership percentages seems to be rather evenly distributed, with no notable skewness. There is much more skewness and spread in the student loan debt data compared to homeownership, but this may be mostly due to the small amount of data available on student loan debt. In general, being white seems to greatly increase one’s chances of owning a home, but the effect on having student loan debt is less clear.


## Median Average Income & Presidents

```{r president, warning = FALSE}
pres_levels <- c("Lyndon B. Johnson", "Richard Nixon", "Gerald Ford", "Jimmy Carter", "Ronald Reagan", "George H. W. Bush", "Bill Clinton", "George W. Bush", "Barack Obama", "Donald Trump")

president <- president %>%
  rename(
    start = "Took office", 
    end = "Left office",
    name = "President"
    ) %>%
  select(c(name, start, end)) %>%
  mutate(start = dmy(start), end = dmy(end)) %>%
  mutate(start = year(start), end = year(end)) %>%
  mutate(end = if_else(is.na(end), 2017, end)) %>%
  add_row(name = "Donald Trump", start = 2017, end = 2021) %>%
  filter(end > 1967) %>%
  mutate(name = fct_relevel(name, pres_levels))

pres_income_df <- income_mean %>%
  filter(race == "All Races",
         income_quintile == "Third") %>%
  group_by(year)

med_income_line <- pres_income_df %>%
  ggplot(mapping = aes(x = year, y = income_dollars)) +
  geom_line() +
  labs(title = "Median Average U.S. Income",
       x = "Year", y = "Average income (2019 dollars)",
       caption = "Source: The Urban Institute") +
  scale_y_continuous(label = scales::dollar)

# this is the mean income of the middle quintile of US

med_income_line + 
  geom_rect(data = president, 
            mapping = aes(NULL, NULL, xmin = start, xmax = end, 
                          ymin = min(pres_income_df$income_dollars), 
                          ymax = max(pres_income_df$income_dollars), 
                          fill = name), 
            alpha = 0.2) +
  labs(fill = "President")   
```

Out of curiosity, we see how the median average income, i.e. the average income for the third income quintile, has changed over different presidencies. We can see that median average income dropped during Carter’s term, likely as part of 1970s stagflation. We also see it initially decreased during the early years of Obama's presidency, likely due to the 2008 financial crisis, but it then returned to 2000 levels before surpassing it as the economy recovered. This graph doesn’t have much explanatory power, as presidents have limited effect on the economy and median average income. Whoever is president doesn’t seem to determine the mean income a family has like race does, as noted earlier. It will be interesting to see how median average income changes when the last year of Donald Trump’s presidency is added, given the economic downturn brought by the coronavirus. 
